tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-23 paper.tex
Progress

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
This success has led to the collection of many population-scale Biobank
datasets~\citep{tanjo2021practical},
but the sampling and resulting vast volumes of data
are presenting significant challenges to GWAS
methology~\citep{uffelmann2021}.

@@

Simulation is a critical component of GWAS method development, and
generally consists of two steps.
Firstly, we must simulate some genetic variation. This is
usually done either by model-based population genetic
simulation using methods such as
\texttt{msprime}~\citep{baumdicker2022efficient}
or \texttt{SLiM}~\citep{haller2023},
or by statistical resampling from existing datasets
using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
Secondly, given this genetic variation dataset, we can
simulate quantitative traits.
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
have been instrumental in making population genetic simulations
of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{wong2023general},
using the open-source \texttt{tskit} library~\citep{ralph2020}.

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store enormous genetic variation datasets compactly. Instead of
storing the genetic sequence variation in the conventional
matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in
gzip-compressed VCF ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss this
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
genealogies~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

>>>>

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been
identified~\citep[e.g.][]{yengo2022,mathieson2023},
% \citep{locke2015,ishigaki2022,
% mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
This success has led to the collection of many population-scale Biobank
datasets~\citep{tanjo2021practical},
but such scale is presenting significant challenges to GWAS
methology~\citep{uffelmann2021}.

@@

Simulation is a critical component of GWAS method development, and
generally consists of first simulating some genetic
variation, and then simulating quantitative traits based on it.
The combined genetic variation and
simulated quantitative traits then represent truth-data
which association testing methods can be evaluated against.
Simulating genetic variation is
usually done either by model-based population genetic
simulation using methods such as
\texttt{msprime}~\citep{baumdicker2022efficient}
or \texttt{SLiM}~\citep{haller2023},
or by statistical resampling from existing datasets
using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
Both approaches have advantages and disadvantages.
Roughly speaking, model-based
simulation methods provide better control of
population processes such as demography but can be overly simplistic,
whereas resampling methods better capture some of the nuance of
real data, while lacking a direct connection to population genetics models.
Model-based population genetic simulations have made
great strides in recent years, with major advances in both
scalability~\citep{kelleher2016efficient,kelleher2018efficient,haller2018tree}
and realism~\citep{adrion2020,anderson2023}, and
have been successfully used to model large-scale GWAS
cohorts~\cite[e.g.][]{martin2017,zaidi2020}.

An important part of these simulation methods is that they
output ARGs using
the ``succinct tree sequence'' encoding~\citep{wong2023general}
and the open-source \texttt{tskit} library~\citep{ralph2020}.
A key benefit of this approach is that it allows us to
store enormous genetic variation datasets compactly.
Instead of storing the genetic sequence variation in the conventional
matrix-based form~\citep[e.g.\ VCF,][]{danecek2011},
we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in
gzip-compressed VCF ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss this
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
genealogies~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
